Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-08-15 Master Thesis/synopsis.tex
Correct characteristic issue in 2.4

\begin{lemma}\label{2.4-pre}
Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field. Let $p=\operatorname{char}(k)$, $\ell\neq p$ be a prime, and $\operatorname{cd}_\ell(k)\leq 2$. Then:
\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^2(k, \mu_{\ell}) \cong {}_\ell\operatorname{Br}(k) \]
\end{lemma}

\begin{proof}
Let $T$ be the absolute inertia group of $K$ and $V\lhd T$ its absolute ramification group. By Theorem \ref{theorem-sylowgroups-inertiagroup}, $V$ is the $p$-Sylow group of $T$ and the $\ell$-Sylow group of $T/V$ is isomorphic to $\mathbb{Z}_\ell$. Therefore, by \cite{Se64} Ch. I, §3 Proposition 15:
\[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
By Theorem \ref{theorem-absoluteinertia}, we have an isomorphism $G_K/T\cong G_k$. Furthermore, we have $\operatorname{cd}_\ell(G_K/T)=\operatorname{cd}_\ell(k)\leq 2$ by requirement,
thus the following Hochschild-Serre spectral sequence degenerates on $E_2$:
\[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
It yields an isomorphism $\mathrm{H}^2(k, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2))$. Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mu_{\ell}$, hence:
\[ \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^1(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T, \mu_{\ell}) \cong \operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell}) \]
By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$ where $\mathcal{C}$ is the class of all finite groups with order prime to $p$. Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mu_{\ell})\cong \mu_{\ell}$. The rest follows from Proposition \ref{prop:kummer-theory}.
\end{proof}

@@

\begin{proof}
Let $L=K(\mu_{\ell},\sqrt{-1})/K$. Consider the restriction and corestriction maps:
\[ \begin{tikzcd}
\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z})\ar[r, "\operatorname{cor}"]& \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) 
\end{tikzcd} \]
We have $\operatorname{cor}\circ\operatorname{res} = [L:K]$. Since $[L:K]$ is coprime to $\ell$, we see that $\operatorname{res}$ is injective. Thus we can assume that $K=L$, i.e. $\mu_{\ell}\subset K$ and $K$ is not formally real. Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2)$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.

@@

Let $v$ be a discrete valuation of $K$ with residue field $k=k(v)$. Consider the following map induced by the restriction maps: 
\[ \phi: \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
where $w$ ranges over all rank-$2$ valuations with $v(w) = v$. By Lemma \ref{rank-dimension-inequality}, we have $\dim(k) = 1$, thus $k$ is a global field of characteristic $0$, hence a number field. \cite{Se64} Ch. II, §4 Prop. 13 shows $\operatorname{cd}_\ell(k)\leq 2$. With Lemma \ref{2.4-pre}, $\phi$ can be identified with
\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
which is injective by the classical Hasse principle. Moreover, its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$. Combining these maps yields the following injection:
\[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof. 
\end{proof}

@@

\begin{proof}
Let $L/K$ be a finite subextension of $\overline{K}/K$. Let $\ell$ and $\ell'$ be the constant fields of $L$ and $L'$ respectively. Applying Lemma \ref{4.2-pre} to $K$ and $L$ gives:
\[ [\ell:k] = [\ell':k'] \]
Therefore $L/K$ is a constant field extension if and only if $L'/K'$ is a constant field extension. Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$. This implies that $\sigma$ induces an isomorphism:
\[G_k \cong \operatorname{Gal}(K\overline{k}/K) \to \operatorname{Gal}(K'\overline{k'}/K')\cong G_{k'}\] 
Everything else follows from Neukirch-Uchida \ref{thm:neukirch-uchida}.
\end{proof}

>>>>

\begin{lemma}\label{2.4-pre}
Let $K$ be henselian with respect to a discrete valuation $v$, $k=k(v)$ its residue field. Let $\ell$ be a prime such that $\ell=\operatorname{char}(k)$ or $\operatorname{cd}_\ell(k)\leq 2$. Then:
\[ \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2)) \cong \mathrm{H}^2(k, \mathbb{Z}/\ell\mathbb{Z}(1)) \cong {}_\ell\operatorname{Br}(k) \]
\end{lemma}

\begin{proof}
Let $p=\operatorname{char}(k)$. If $\ell = p$, the first isomorphism follows from \cite{Ka86} Lemma 1.4 (3) and the second one from \cite{Ka86} §0. Now assume $\ell\neq p$ and $\operatorname{cd}_\ell(k)\leq 2$.

Let $T$ be the absolute inertia group of $K$ and $V\lhd T$ its absolute ramification group. By Theorem \ref{theorem-sylowgroups-inertiagroup}, $V$ is the $p$-Sylow group of $T$ and the $\ell$-Sylow group of $T/V$ is isomorphic to $\mathbb{Z}_\ell$. Therefore, by \cite{Se64} Ch. I, §3 Proposition 15:
\[ \operatorname{cd}_\ell(T)\leq \operatorname{cd}_\ell(V) + \operatorname{cd}_\ell(\mathbb{Z}_\ell) = 1 \]
By Theorem \ref{theorem-absoluteinertia}, we have an isomorphism $G_K/T\cong G_k$. Furthermore, we have $\operatorname{cd}_\ell(G_K/T)=\operatorname{cd}_\ell(k)\leq 2$ by requirement,
thus the following Hochschild-Serre spectral sequence degenerates on $E_2$:
\[E_2^{i,j}=\mathrm{H}^i(G_K/T, \mathrm{H}^j(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\Longrightarrow\mathrm{H}^{i+j}(G_K, \mathbb{Z}/\ell\mathbb{Z}(2))\]
It yields an isomorphism $\mathrm{H}^2(k, \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)))\cong \mathrm{H}^3(K, \mathbb{Z}/\ell\mathbb{Z}(2))$. Since $\mu_{\ell}$ is contained in the absolute inertia field of $K$, $T$ acts trivially on $\mathbb{Z}/\ell\mathbb{Z}(1)= \mu_{\ell}$ and $\mathbb{Z}/\ell\mathbb{Z}(2)=\mu_{\ell}^{\otimes 2}$, hence:
\begin{align*}
\mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(2)) &\cong \mathrm{H}^1(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
&\cong \operatorname{Hom}_\text{cont.}(T, \mathbb{Z}/\ell\mathbb{Z}(1)) \\
&\cong \operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))
\end{align*}
By Theorem \ref{theorem-sylowgroups-inertiagroup}, $T/V$ is the free pro-$\mathcal{C}$ group of rank $1$ where $\mathcal{C}$ is the class of all finite groups with order prime to $p$. Therefore, $\operatorname{Hom}_\text{cont.}(T/V, \mathbb{Z}/\ell\mathbb{Z}(1))\cong \mathbb{Z}/\ell\mathbb{Z}(1)$. The second isomorphism follows from Proposition \ref{prop:kummer-theory}
\end{proof}

@@

\begin{proof}
Let $L=K(\mu_{\ell},\sqrt{-1})/K$. Consider the restriction and corestriction maps:
\[ \begin{tikzcd}
\mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(L,\mathbb{Z}/\ell\mathbb{Z})\ar[r, "\operatorname{cor}"]& \mathrm{H}^3(K,\mathbb{Z}/\ell\mathbb{Z}) 
\end{tikzcd} \]
We have $\operatorname{cor}\circ\operatorname{res} = [L:K]$. Since $[L:K]$ is coprime to $\ell$, we see that $\operatorname{res}$ is injective. Thus we can assume that $K=L$, i.e. $\mu_{\ell}\subset K$ and $K$ is not formally real. Now we can replace $\mathbb{Z}/\ell\mathbb{Z}$ by $\mathbb{Z}/\ell\mathbb{Z}(2) = \mu_{\ell}^{\otimes 2}$ and let us write $\mathrm{H}^3(L) = \mathrm{H}^3(L, \mathbb{Z}/\ell\mathbb{Z}(2))$.

@@

Let $v$ be a discrete valuation of $K$ with residue field $k=k(v)$. Consider the following map induced by the restriction maps: 
\[ \phi: \mathrm{H}^3(K_v) \longrightarrow \prod_w\mathrm{H}^3(K_w) \]
where $w$ ranges over all rank-$2$ valuations with $v(w) = v$. By Lemma \ref{rank-dimension-inequality}, we have $\dim(k) = 1$, thus $k$ is a global field. \cite{Se64} Ch. II, §4 Prop. 11 and \cite{Se64} Ch. II, §4 Prop. 13 show $\operatorname{cd}_\ell(k)\leq 2$. With Lemma \ref{2.4-pre}, $\phi$ can be identified with
\[ {}_\ell\operatorname{Br}(k) \longrightarrow \prod_{\widetilde{w}\in\operatorname{Spv}^{(1)}(k)} {}_\ell\operatorname{Br}(k_{\widetilde{w}}) \]
which is injective by the classical Hasse principle. Moreover, its image is contained in $\bigoplus {}_\ell\operatorname{Br}(k_{\widetilde{w}})$. Combining these maps yields the following injection:
\[\Phi: \mathrm{H}^3(K)\longrightarrow \bigoplus_{x\in\mathfrak{X}_1}\  \bigoplus_{\substack{w\in\operatorname{Spv}_\text{rd}^{(2)}(K) \\ v(w) = v_x}} \mathrm{H}^3(K_w) \]
The direct sum ranges over a subset of $|\operatorname{Spv}(K)|$, thus this concludes the proof. 
\end{proof}

@@

\begin{proof}
Let $L/K$ be a finite subextension of $\overline{K}/K$. Let $\ell$ and $\ell'$ be the constant fields of $L$ and $L'$ respectively. Applying Lemma \ref{4.2-pre} to $K$ and $L$ gives:
\[ [\ell:k] = [\ell':k'] \]
Therefore, $L/K$ is a constant field extension if and only if $L'/K'$ is a constant field extension. Since $K\overline{k}/K$ and $K'\overline{k'}/K$ are the union of their finite subextensions, we obtain $(K\overline{k})' = K'\overline{k'}$. This implies that $\sigma$ induces an isomorphism:
\[G_k \cong \operatorname{Gal}(K\overline{k}/K) \to \operatorname{Gal}(K'\overline{k'}/K')\cong G_{k'}\] 
Everything else follows from Neukirch-Uchida \ref{thm:neukirch-uchida}.
\end{proof}
